\texttt{Taxonomic supertree construction with}\texttt{\emph{ Incertae sedis}}\texttt{ taxa}

### 2018-04-26 doc/incertae_sedis.tex
Update.

\subsection{Background}

\paragraph{Previous algorithm}

@@

\note{     1 Details & Methods

1.1 Some examples in OTT This paragraph pulled from intro to
otcetera/doc /handling-incertae-sedis.pdf

The problem is becoming more acute because NCBI is putting a larger
number of taxa into groups that are marked as ``classified''. or
example, when OTT 2.9 was created NCBI's clasification of the bird
family Sylvidae included a group five genera that were placed within
``[http://www.ncbi.nlm.nih.gov/Taxonomy/Brows er/wwwtax.cgi?mode=Undef&
id=36270&lvl=3&lin=f&keep=1&srchmode=1&unlock||unclassi fied
Sylviidae].'' This includes the genus Regulus.
Thus in OTT 2.9 Regulus
(ott ID = 3599326) is placed inside the family Silvidae (OTT ID =
259942).
The ``unclassified Sylviidae'' does not appear in OTT; instead
Regulus is flagged as ``unclassified,sibling higher'' and all of the
species within Regulus are flagged as ``unclassified inherited.''

The taxonomy (as of version 2.9) also contains 685 cases of taxa that
are flagged as both ``unclassified inherited'' and ``unclassified''based
on grep unclass.*unclassified\_inherited taxonomy.tsv.

While we get Regulus back, we do not end up placing it in Sylviidae
because Sylviidae is broken.
It conflicts with about 20 different
trees.

1.2 Taxonomy details

Flags.
The Open Tree's reference taxonomy is produced by a tool called
smasher that notices hints that a taxon is incertae sedis and labels
that taxon with one of five flagssee https://github.com/OpenTreeOfLife
/reference-taxonomy/wiki /Taxon-flags.This document is not concerned
with the flagging system per se, so ``incertae sedis'' will be used here
to refer to all of the flags that denote taxa with uncertain
placement..
While a taxonomist may intend to use incertae sedis to
indicate a limited number of possible positions for a taxon to go in
the taxonomy, we do not retain any such details.
Container nodes.
Some taxonomies have nodes with names like ``Incertae
sedis (Bacteria)''.
Such nodes are called containers.
They indicate
that each child node is an incertae sedis child of the container
node's parent.
The OTT retains container nodes with the was_container
flag, but seems to have already moved all of the container's children
to the container's parent.
}
\begin{figure}
\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-clade/tree1-shaded}

@@

\section{Synthesis and conflict resolution with incertae sedis taxa}

@@

>>>>

\subsection{Background}
\label{sec:background}
\paragraph{Previous algorithm}

@@

\begin{figure}
\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-clade/tree1-shaded}

@@

\section{The rooted-split semantics in practice}

In order to examine the results of the rooted splits-semantics above, we explore
the outcome on a series of complex cases.  These cases are intended to
illustrate some of the issues that arise when developing a semantics of
\emph{incertae sedis}, as well as the outcomes for the rooted-split semantics.
We think that they constitute one of the main contributions of the paper.  In
each case, the synthesis trees in this section are constructed using the
BUILD-based algorithm described in section \ref{sec:background}, without
decomposing them into smaller subproblems or pruning and then unpruning
taxonomy-only taxa.  Synthesis trees that are described as \emph{incertae
sedis}-aware apply the BUILD algorithm to splits constructed using equation
\ref{eq:exsetformonea}, whereas the synthesis trees that are not \emph{incertae
sedis}-aware are constructed from splits using equation
\ref{eq:exsetformonea-traditional}.

Much of the complexity that we explore here comes from the fact that it is not
only tip taxa that can be \emph{incertae sedis}.  It is also possible to have
higher ranked  \emph{incertae sedis} taxa, such as genera that are marked
\emph{incertae familia}.

\subsection{The \emph{incertae sedis} semantics restores taxonomic groups and names}

With the naive semantics, intrusion of an incertae sedis taxon into a sibling is
interpreted as conflict with that sibling.  This results in the loss of the edge
corresponding to the sibling taxon as well as the taxon name.

We see from Fig. \ref{fig:Handling-incertae-sedis}d that use of the incertae sedis
semantics results the taxonomy edge $AB$ grouping $A$ and $B$.  It also restores the
names $AB$ and $A$.  We may thus determine that $C$ has been placed within $A$ (and
its ancestor $AB$).

@@

\section{Details \& Methods}

\subsection{Cases of incertae sedis taxa}
1.1 Some examples in OTT This paragraph pulled from intro to
otcetera/doc /handling-incertae-sedis.pdf

The problem is becoming more acute because NCBI is putting a larger
number of taxa into groups that are marked as ``classified''. For
example, when OTT 2.9 was created NCBI's clasification of the bird
family Sylvidae included a group five genera that were placed within
\url{http://www.ncbi.nlm.nih.gov/Taxonomy/Browser/wwwtax.cgi?mode=Undef&id=36270&lvl=3&lin=f&keep=1&srchmode=1&unlock||unclassified Sylviidae}.
This includes the genus Regulus. Thus in OTT 2.9 Regulus (ott ID = 3599326) is placed inside the family Silvidae (OTT ID =
259942).
The ``unclassified Sylviidae'' does not appear in OTT; instead
Regulus is flagged as ``unclassified,sibling higher'' and all of the
species within Regulus are flagged as ``unclassified inherited.''

The taxonomy (as of version 2.9) also contains 685 cases of taxa that
are flagged as both ``unclassified inherited'' and ``unclassified'' based
on {\tt grep unclass.*unclassified\_inherited taxonomy.tsv}.

\bdrcomment{We get Regulus back, bu we do not end up placing it in Sylviidae
because Sylviidae is broken. It conflicts with about 20 different trees.}

\subsection{Taxonomy details}

\paragraph{Flags.}
The Open Tree's reference taxonomy is produced by a tool called
smasher that notices hints that a taxon is incertae sedis and labels
that taxon with one of five flags (see https://github.com/OpenTreeOfLife
/reference-taxonomy/wiki /Taxon-flags). This document is not concerned
with the flagging system per se, so ``incertae sedis'' will be used here
to refer to all of the flags that denote taxa with uncertain
placement.

\paragraph{Limited attachment range}
While a taxonomist may intend to use incertae sedis to
indicate a limited number of possible positions for a taxon to go in
the taxonomy, we do not retain any such details.

\paragraph{Container nodes.}
Some taxonomies have nodes with names like ``Incertae sedis (Bacteria)''.
Such nodes are called containers. They indicate that each child node is an
incertae sedis child of the container node's parent. The OTT retains container
nodes with the was\_container flag, but seems to have already moved all of the
container's children to the container's parent.
